---
title: "Schedule"
---

The class will cover a series of modules, each of which may be covered in more than or less than a single class period.  My notes for these modules will be updated regularly [here](https://kevinlanning.github.io/DataSciLibArts/).

```{r echo=FALSE, message=FALSE}
library(knitr)
library(googlesheets)
library(kableExtra)
suppressMessages(library(dplyr))
sched <- gs_title("DataSciSpring2018Schedule") %>% #  register a googlesheet
    gs_read(ws=1) # and read in the first worksheet as a data table
#    gs_read_cellfeed(ws=1) to read in formulas as well as contents
sched[is.na(sched)] <-"tba"

```


## Introduction (weeks 1-2)

```{r echo=FALSE, message=FALSE}
kable(sched[c(1:2,4),c(1,3:7)], "html") %>%
    kable_styling(bootstrap_options = c("striped", 
                                        "hover", "condensed", "responsive"))

```

*1 Data science for the liberal arts

2 Setting up your computer

3 Pretest

4 Some basic tools: Slack, Markdown, Google Docs

5 An introduction to R 

6 Finding help

## Towards data literacy (wks 3-5)
```{r echo=FALSE, message=FALSE}
kable(sched[5:9,c(1,3:7)], "html") %>%
    kable_styling(bootstrap_options = c("striped", 
                                        "hover", "condensed", "responsive"))

```

7 Empiricism > intuition

8 The ubiquity of data

9 Counting and observing

10 Principles of data visualization 

11 Probability

12 Inference

13 Reproducibility

## Towards data proficiency (wks 5-9)
```{r echo=FALSE, message=FALSE}
kable(sched[9:15,c(1,3:7)], "html") %>%
    kable_styling(bootstrap_options = c("striped", 
                                        "hover", "condensed", "responsive"))

```

14 Literate programming with R markdown 

15 Data visualization in R with ggplot2

16 Exploratory Data Analysis

17 Tidy data and the tidyverse

18 Messy data: Cleaning and curation

19 Selecting data in R with dplyr

20 Manipulating data in R with dplyr

21 Working with relational data

22 Working with strings, factors, dates, and times

23 Working with lists 

24 Open science

## Towards data fluency (wks 9-13)
```{r echo=FALSE, message=FALSE}
kable(sched[c(16,19:27),c(1,3:7)], "html") %>%
    kable_styling(bootstrap_options = c("striped", 
                                        "hover", "condensed", "responsive"))

```

25 Writing functions

26 Writing loops

27 Getting data: Web scraping and APIs 

28 Linear and multiple regression

29 Logistic regression

30 Classification and types of errors

31 Cross-validation

32 Some machine learning techniques

33 Concepts in computational linguistics

34 Analyzing text data

35 From graphs to networks

36 Empirical findings in network analysis

37 Analyzing networks 

38 Categories and classification

39 Making interactive visualizations with R Shiny

## Towards data leadership (wks 13-15)
```{r echo=FALSE, message=FALSE}
kable(sched[28:32,c(1,3:7)], "html") %>%
    kable_styling(bootstrap_options = c("striped", 
                                        "hover", "condensed", "responsive"))

```

40 The significance of scalability 

41 Digital humanities and data science

42 Humility and teamwork

43 Ethics and data responsibility

